Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-16 paper/intro.tex
Moved realted work
Wordsmithing intro

Introductory programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 60s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below. 
\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    \stepcounter{enumi}
  \item[] \labeltext[\textsc{d}2]{}{wp-assign}
    \begin{enumerate}[leftmargin=0pt]
    \item \label{wp-write}
      $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    \item \label{wp-let}
      $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    \item \label{wp-read}
      $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
    \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \\
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.  We have split
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
\eqref{wp-assign} into three cases. In our notation,
$\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
most once,
$\aLoc$--$\cLoc$ range over shared memory references, and
$\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
include
$\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
  equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}

This is quite a pretty explanation of sequential computation in a sequential
context.  In a concurrent context, however, \ref{wp-read} is unsound!
\ref{wp-read} assumes that a read from
$x$ must be fulfilled by a preceding write to
$x$.  In a concurrent context, writes may come from other threads.

Existing approaches to sequential composition in the concurrent context
either assume exclusive access, as in concurrent separation logic
\cite{OHearn:2007:RCL:1235896.1236121}, or abandon the logical approach
altogether, as in the pomset model of
\citet{DBLP:journals/corr/abs-1804-04214}---this model uses syntactic
dependencies and thus dramatically limits compiler optimization.  This leaves
open the question of how to apply logic to racy programs without
overstraining the implementation.  To understand the solution, one must first
understand the constraints imposed by hardware and compilers.

@@

Pomsets with Preconditions show how the logical approach to sequential
dependency calculation can be mixed into a relaxed memory model.  However,
\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} do not provide a model of
sequential composition.  Instead, their model uses \emph{prefixing}, which
requires that the model is built from right to left: events are prepended one
at a time, with perfect knowledge of the future.  This makes reasoning about
sequential program fragments difficult.  For example,
\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} state the equivalence
allowing reordering independent writes as follows,
\begin{align*}
  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp\SEMI\aCmd} &=
  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp\SEMI\aCmd} \;\;\text{if}\; \aLoc\neq\bLoc
\end{align*}
where $\aCmd$ is the entire future computation!  By formalizing sequential
composition, we can show:
\begin{align*}
  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
\end{align*}
Then the equivalence holds in any (sequential) context.

@@

\section{Related Work}

>>>>

Introductory programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 1960s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$ ranges over thread-local \emph{registers}
and $\aExp$ ranges over side-effect free \emph{expressions}). 
\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    \stepcounter{enumi}
  %\item[] \labeltext[\textsc{d}2]{}{wp-assign}
  %  \begin{enumerate}[leftmargin=0pt]
  %  \item \label{wp-write}
  %    $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    \item \label{wp-let}
      $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
  %  \item \label{wp-read}
  %    $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
  %  \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \\
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.
%% We have split
%% \citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
%% \eqref{wp-assign} into three cases. In our notation,
%% $\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
%% most once,
%% $\aLoc$--$\cLoc$ range over shared memory references, and
%% $\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
%% include
%% $\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
%%   equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}
%%
This is an elegant explanation of sequential computation in a sequential
context.  In a concurrent context, however, generalizing
\ref{wp-let} from thread-local registers to cross-thread memory locations
is unsound! \ref{wp-let} is sound because a read from a thread-local
register must be fulfilled by a preceding write
in the same thread.  In a concurrent context, aread from a cross-thread
memory location may be fulfilled by a write
in another thread.

Existing approaches to sequential composition in the concurrent context
either assume exclusive access, as in concurrent separation logic
\cite{OHearn:2007:RCL:1235896.1236121}, or abandon the logical approach
altogether, as in the pomset model of
\citet{DBLP:journals/corr/abs-1804-04214}---this model uses syntactic
dependencies and thus dramatically limits compiler optimization.  This leaves
open the question of how to apply logic to racy programs without
overconstraining the implementation.  To understand the solution, one must first
understand the constraints imposed by hardware and compilers.

@@

Pomsets with Preconditions show how the logical approach to sequential
dependency calculation can be mixed into a relaxed memory model.  However,
\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} do not provide a model of
sequential composition.  Instead, their model uses \emph{prefixing}, which
requires that the model is built from right to left: events are prepended one
at a time, with perfect knowledge of the future.  This makes reasoning about
sequential program fragments difficult.  For example,
\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} state the equivalence
allowing reordering independent writes as follows,
\begin{align*}
  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp\SEMI\aCmd} &=
  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp\SEMI\aCmd} \;\;\text{if}\; \aLoc\neq\bLoc
\end{align*}
where $\aCmd$ is the entire future computation!  By formalizing sequential
composition, we can show:
\begin{align*}
  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
\end{align*}
Then the equivalence holds in any
%% (sequential)
context.

@@

\subsection{Related Work}
